\section{Attempts on defining equivalence between timed and untimed semantics}

Two utbs may still have the same observable behavior, but
differ in timers that are started but never time out. We will introduce two
ways to consider this.
For a sequence $i_1/o_1 \ldots i_j/o_j$ of input-output pairs and
$l \leq j$, say that $y_l$ is \emph{live after $i_1/o_1 \ldots i_j/o_j$} in $\M$
if there is an utb of $\M$ of form
${i_1/o_1/\varrho_1}  \cdots {i_j/o_j/\varrho_j} \cdots {\toevent{y_l}/o_k/\varrho_k}$
with $j < k$,  i.e., which extends the sequence
$i_1/o_1 \ldots i_j/o_j$ so that $y_l$ expires at some later transition.

For a utb $\beta  =  {i_1/o_1/\varrho_1}  \cdots {i_k/o_k/\varrho_k}$ of $\M$,
let $\timerlive{\M}{\beta}$ be the utb obtained by removing 
$y_l$ from $\domof{\varrho_j}$ whenever $0 < l \leq j \leq k$ and
$y_l$ is not live after $i_1/o_1 \ldots i_j/o_j$.
Let
$\timerlivebehs{\M}$ be $\setcomp{\timerlive{\M}{\beta}}{\beta \in \can{\M}}$.
We say that $\M \approx_{\mathit{untimed}} \N$ if
$\timerlivebehs{\M} = \timerlivebehs{\N}$.


If $\beta  =  {i_1/o_1/\varrho_1}  \cdots {i_k/o_k/\varrho_k} \in \timerlivebehs{\M}$,
we say that a timer $y_l \in \domof{\varrho_k}$
is {\em expirable} after $\beta$ if $\beta \toevent{y_l}/o/\rho \in \timerlivebehs{\M}$ for some $o$ and $\rho$.
The following lemma characterizes when this can happen.

\begin{lemma}
\label{expirable}
Let $\beta  =  {i_1/o_1/\varrho_1}  \cdots {i_k/o_k/\varrho_k} \in \timerlivebehs{\M}$.
Then $y_l$ is expirable after $\beta$ iff 
there is a solution $t_0, \ldots , t_{k+1}$ of  $\Constraints{\beta}$ such that
$\varrho_l(y_l) - (t_k - t_l)$ is minimal among all
$\varrho_j(y_j) - (t_k - t_j)$ with $y_j \in \domof{\varrho_k}$.
\end{lemma}
\begin{proof}
Follows from Lemma~\ref{lem:timerval}, which states that
$\tvals_k(y_j) = \varrho_j(y_j) - (t_k - t_j)$ for each $y_j \in \domof{\varrho_k}$.
\end{proof}

We are now in the position to characterize the set of
timed words of $\M$ in terms of the set $\timerlivebehs{\M}$.

\begin{theorem}
\label{thm:correspondence}
$w =  d_1 ~ i_1' ~ o_1 \cdots d_k ~ i_k' ~ o_k$
  is a timed word of $\M$ if and only if there is a utb 
  $\beta = {i_1/o_1/\varrho_1}  \cdots {i_k/o_k/\varrho_k}$
   in $\timerlivebehs{\M}$,
   such that (letting $t_j = d_1 + \cdots + d_j$)
   $t_0, \ldots, t_k$ satisfies $\Constraints{\beta}$, and
  such that 
$i'_j   =   i_j$ if $i_j \in I$ and
  $i'_j   = \mathit{to}$ if $i_j \in \toeventsof{Y}$
  for $j = 1 , \ldots , k$.
\end{theorem}
  
\begin{proof}
First assume that
$w =  d_1 ~ i_1' ~ o_1  \cdots d_k ~ i_k' ~ o_k$
is a timed word of $\M$, i.e., there is a timed run of $\M$
\begin{eqnarray*}
\alpha & = & (q_0,\tvals_0) \xrightarrow{d_1} (q_0,\tvals_0') \xrightarrow{i_1/o_1/\rho_1} (q_1,\tvals_1) \xrightarrow{d_2} 
%C'_1 \xrightarrow{i_2/o_2} C_2 
\cdots
\\ && \qquad \cdots
\xrightarrow{d_k} (q_{k-1},\tvals_{k-1}') \xrightarrow{i_k/o_k/\rho_k} (q_k,\tvals_k)
\end{eqnarray*}
where for all  $j$,
$i'_j   =   i_j$ if $i_j \in I$, and
$i'_j   = \mathit{to}$ if $i_j \in \toevents$.
Let $\beta = \timerlive{\M}{\can{\alpha}}$.
Let $t_j = d_1 + \cdots + d_j$ for $j = 1 , \ldots, k+1$.
We must prove that $t_0, \ldots t_k$ is a solution to $\Constraints{\beta}$.

Let $y_l \in Y$,
%% i.e., be a timer that is started in the $l$th transition, i.e.,
%% $\rho_l(x_p) \in \natplus$.
Using the definition of $\can{\alpha}$, we establish by induction that 
 whenever $l < j \leq k$ and
 $y_l \in \domof{\varrho_{j-1}}$, then
\begin{eqnarray*}
 \tvals_{j-1}'(\invcanmap{\alpha}{j-1}(y_l)) & = &
 \varrho_l(y_l) - (d_{l+1} + \cdots + d_j)
 \ \ .
\end{eqnarray*}
 For the base case $j = l+1$, let $x_p = \invcanmap{\alpha}{l}(y_l)$.
 By the definition of $\canmap{\alpha}{l}$ we have that
 $\rho_l(x_p)  \in \natplus$ and  
 $\varrho_l(y_l) = \rho_l(x_p)$. By the semantics of timed runs,
 $\tvals_l(x_p) =  \rho_l(x_p) = \varrho_l(y_l)$ and
 $\tvals_l'(x_p) =  \varrho_l(y_l) - d_{l+1}$.
 For the inductive step, assume that $l+1 < j$ and 
 $y_l \in \domof{\varrho_{j-1}}$, and
 $\tvals_{j-2}'(\invcanmap{\alpha}{j-2}(y_l)) =
 \varrho_l(y_l) - (d_{l+1} + \cdots + d_{j-1})$.
 Let $x_p = \invcanmap{\alpha}{j-1}(y_l)$.
 Then $\rho_{j-1}(x_p) \not\in \natplus$, whence
 $\tvals_{j-1}'(x_p) =  \tvals_{j-1}(x_p) -d_j = 
 \tvals_{j-2}'(\rho_{j-1}(x_p)) - d_j$.
 From the definition of $\varrho_{j-1}$, we infer
 $\rho_{j-1}(x_p) = \rho_{j-1}(\invcanmap{\alpha}{j-1}(y_l)) =
 \invcanmap{\alpha}{j-2}(\varrho_{j-1}(y_l)) =
 \invcanmap{\alpha}{j-2}(y_2)$. Using the inductive hypothesis, we infer
 $\tvals_{j-1}'(\invcanmap{\alpha}{j-1}(y_l)) =
 \varrho_l(y_l) - (d_{l+1} + \cdots + d_j)$, which proves the inductive
 step.

 Let us now consider the two types of constraints in $\Constraints{\beta}$.
 \begin{itemize}
 \item $l \sdelay{\beta} j$. This means that the $j$th input in $\beta$ is
   $\toevent{y_l}$. By the semantics of timed runs, we must then have
   $\tvals_{j-1}'(\invcanmap{\alpha}{j-1}(y_l)) = 0$ By Property XXX, this means
   $\varrho_l(y_l) = (d_{l+1} + \cdots + d_j) = t_j - t_l$, i.e.,
   $t_0, \ldots, t_k$ satisfies the corresponding constraint.
 \item $l \wdelay{\beta} j$. This means that $\invcanmap{\alpha}{j-1}(y_l)$
   is defined, 
   By the semantics of timed runs, we must then have
   $\tvals_{j-1}'(\invcanmap{\alpha}{j-1}(y_l)) \geq 0$.
   By Property XXX, this means
   $\varrho_l(y_l) \geq (d_{l+1} + \cdots + d_j) = t_j - t_l$, i.e.,
   $t_0, \ldots, t_k$ satisfies the corresponding constraint.
 \end{itemize}
In the other direction, assume that 
 $\beta = {i_1/o_1/\varrho_1}  \cdots {i_k/o_k/\varrho_k}$
   in $\timerlivebehs{\M}$,
and that
$t_0, \ldots, t_k$ satisfies $\Constraints{\beta}$.
We will then prove that
\begin{eqnarray*}
\alpha & = & (q_0,\tvals_0) \xrightarrow{d_1} (q_0,\tvals_0') \xrightarrow{i_1/o_1/\rho_1} (q_1,\tvals_1) \xrightarrow{d_2} 
%C'_1 \xrightarrow{i_2/o_2} C_2 
\cdots
\\ && \qquad \cdots
\xrightarrow{d_k} (q_{k-1},\tvals_{k-1}') \xrightarrow{i_k/o_k/\rho_k} (q_k,\tvals_k)
\end{eqnarray*}
is a timed run. We must establish that $\alpha$
satisfies the requirements of a timed run. This is done
using property XXX, by the same arguments
as the first direction of the proof. We must only check that timers that
are not in $\invcanmap{\alpha}{j}(\domof{\varrho_j})$ for some $j$ do not
interfere. This can happen only if such a timer blocks a delay transition,
by being smaller than any clock of form
$\invcanmap{\alpha}{l}(\domof{\varrho_l})$.
But this implies that there is also a timed run where such a timer expires,
which implies that it should not have been removed when forming
$\timerlivebehs{\M}$.
\end{proof}

\section{Potentially useful Text Snippets}

\tofix{Do not read this Subsection}

Let $\lean{\beta}$ be the utb obtained by removing 
$y_l$ from all $\beta$ altogether (i.e., from each $\domof{\varrho_j}$)
whenever $y_l$ does not expire in $\beta$.
Let $\can{\M}$ be the set of utbs of $\M$,, and
let 
$\lean{S}$ be $\setcomp{\lean{\beta}}{\beta \in S}$.
For a lean utb $\beta \in S$
\begin{eqnarray*}
\beta & = & X_0 \xrightarrow{i_1/o_1/\rho_1} X_1  \cdots \xrightarrow{i_k/o_k/\rho_k} X_{k},
\end{eqnarray*}
define its \emph{saturation wrp.\ $S$}, denoted $\saturatebeh{\beta}{S}$,
as the utb obtained from $\beta$ by performing the operation
$\Addtimerop{l}{d}{j}$ 
for each $\beta' \in S$ and $l < j \leq k$, such that $\beta'$ has the same
$j$ first input and outputs as $\beta$, which sets $x_l$ to $d$ and does not stop
or expire $x_l$ up to and including the $j$th transition (i.e., $\beta'$ is of
form
\begin{eqnarray*}
\beta' & = & X_0 \xrightarrow{i_1/o_1/\rho_1'} \cdots \xrightarrow{i_j/o_j/\rho_j'} X_{j}' \xrightarrow{i'/o'/\rho'}  \cdots
\end{eqnarray*}
with $\rho_l'(x_l) = d$, and $x_l \in X_j'$).
For a lean timer language $S$ define its \emph{saturation} $\saturateset{S}$
as $\setcomp{\saturatebeh{\beta}{S}}{\beta \in S}$.

Consider a utb $\beta$
\begin{eqnarray*}
\beta & = & Y_0 \xrightarrow{i_1/o_1, \rho_1} X_1  \cdots \xrightarrow{i_k/o_k, \rho_k} Y_{k}.
\end{eqnarray*}
Let $1 \leq l \leq j \leq k$, $d \in \natplus$ and suppose that 
(1) $\rho_l$ either starts no timer or sets $y_l$ to $d$,
(2) $\beta$ does not contain a timeout event $\toevent{y_l}$.
Then $\Addtimer{l}{d}{j}{\beta}$ is the utb obtained from $\beta$ by
letting $\rho_l$ map $y_l$ to $d$, and adding $y_j$ to the sets $Y_l$ up to (and including) $Y_j$.
%In the utb $\emptyset \xrightarrow{i/o,~ x:=1 } \{ x\} \xrightarrow{i/o,~ y:=60 } \{ x, y\}$, for instance, we may remove timer $x$
%and obtain $\emptyset \xrightarrow{i/o } \emptyset \xrightarrow{i/o,~ y:=60 } \{ y\}$.
%Similarly, we may remove the first use of timer $x$ in the utb 
%$\emptyset \xrightarrow{i/o,~ x:=1 } \{ x\} \xrightarrow{i/o,~ x:=5 } \{ x\} \xrightarrow{\toevent{x}/o} \emptyset$ and obtain
%$\emptyset \xrightarrow{i/o } \emptyset \xrightarrow{i/o,~ x:=5 } \{ x\} \xrightarrow{\toevent{x}/o} \emptyset$.
Write $\beta \sqsubseteq \beta'$ if $\beta'$ can be obtained from $\beta$ by zero or more applications of function $\mathsf{addtimer}.$
Then $\sqsubseteq$ is a partial order whose minimal elements are utbs in which every timer that is started also times out.
We call such utbs \emph{lean}.

\todobj{Maybe scrap the following}
We say that $\beta$ is a \emph{partial} utb of MMT $\M$ iff $\M$ has an utb $\beta'$ such that $\beta \sqsubseteq \beta'$.


The set of feasible lean behaviors of an MMT has certain properties, that
we here collect into the definition of a \emph{timer language}.

\begin{definition}
\label{def:timer language}
A (lean) \emph{timer language} over $I$ and $O$ is a nonempty set 
$S$ of (feasible?) lean utbs in canonical form over $I$ and $O$ that satisfies the following five properties:
\begin{itemize}
\item
\emph{no initial timers}: $\beta \in S \Rightarrow \Head{\beta} = \emptyset$,
\item
\emph{prefix closed}: $\beta \cdot \gamma \in S \Rightarrow \beta \in S$,
\item
\emph{behavior deterministic}:
$\beta \xrightarrow{i/o_1/\rho_1} X_1 \in S \wedge \beta \xrightarrow{i/o_2/\rho_2} X_2 \in S \Rightarrow o_1 = o_2 \wedge \rho_1 = \rho_2 \wedge X_1 = X_2$,
%% \item
%% \emph{input complete}:
%% $\beta \in S \wedge i \in I \Rightarrow \exists o, \rho, Y : \beta \xrightarrow{i/o/\rho} Y \in S$,
\item
\emph{timer live}:
$\beta \in S \wedge x \in last(S) \Rightarrow \exists \gamma, o, \rho, Y : \beta \cdot \gamma \xrightarrow{\toevent{x}/o/\rho} Y \in S$,
and
\item
\emph{timeout complete}:
$\beta \in S \wedge x \mbox{ expirable after } \beta \Rightarrow
\exists o, \rho, Y: \beta \xrightarrow{\toevent{x}/o/\rho} Y \in S$.
\end{itemize}
\end{definition}
\todobj{Check again the above definition!!!}

From the saturated version of $\Constraints{\beta}$, we also derive a constraint,
denoted $\Zone{\beta}$,
which characterizes the possible timer valuations $\tvals_k$ in a timed run
$\alpha$ with $\can{\alpha} = \beta$.
The constraint $\Zone{\beta}$
contains for each pair of timers $x_i,x_j$ in $\domof{\rho_k}$ the conjunct
\[
(m + \rho_j(x_j) - \rho_i(x_i)) < \tvals_k(x_j) -\tvals_k(x_i) < (n + \rho_j(x_j) - \rho_i(x_i))
\]
whenever the saturated version of $\Constraints{\beta}$ contains the conjunct
\(
m < t_j - t_i < n
\).
(This can be derived using
$\tvals_k(x_j) = \rho_j(x_j) - (t_k - t_j)$
and
$\tvals_k(x_i) = \rho_i(x_i) - (t_k - t_i)$).

Let $\beta$ be a feasible untimed behavior and let $x \in X$ be a timer. Then we say that $x$ is \emph{expirable} after $\beta$
if there exists a valuation in $\Zone{\beta}$ in which $x$ is minimal.



\section{Another Try on the Precise Nerode Equivalence}

Let us introduce notation to make the subsequent development more convenient. 
Recall that a canonical untimed behavior is \emph{canonical} when, for each $j$,
the timer that is updated in the $j$-th event (if any) is equal to $x_j$.
It is {\em lean} if it is canonical and
includes only timers that expire during the behavior. 
Since timers in lean behaviors are uniquely determined by the labels on
transition, we can denote lean behaviors simply as sequences
$i_1/o_1,\rho_1 \cdots i_n/o_n,\rho_n$ of input/output/assignment triples.
For behaviors that are to be appended as suffixes, we need to distinguish
timers that are assigned in the suffix from timers that are assigned in a prefix.
Therefore, extend the set of timers by the set $Y = \set{y_1,y_2,\ldots}$ of
{\em suffix timers}, which is disjoint from $X$.
Define a {\em suffix behavior} to be a sequence
$i_1/o_1,\rho_1 \cdots i_m/o_m,\rho_m$ of input/output/assignment triples,
in which the inputs are in $I$, of form $\toevent{x_j}$ for $x_j \in X$, or
of form $\toevent{y_j}$ for $y_j \in Y$, but where assignments may assign only
to timers in $Y$. Such a suffix behavior is {\em lean} if
timer $y_j$ is assigned only by $\rho_j$, each timeout event occurs at most once, and all assigned timers in $Y$ expire sometime after their assignment.

For integer $k \geq 0$, let $\suffmap{k}$ be the injective mapping on
$Y$ which maps each $y_j$ to $y_{j+k}$.  We apply mappings $\suffmap{k}$ to
lean suffix behaviors in the natural way.

Let $\beta$ be a lean behavior. We say that 
a lean suffix behavior $\gamma$ is a {\em $\beta$-suffix} if
there is a canonical behavior $\beta'$ with $\beta \sqsubseteq \beta'$ such that
$\beta'\cdot\suffmap{|\beta|}(\gamma)$ is a lean behavior.

We can define general residual languages, by letting
$\suffixbehs{S}{\beta}$ be the set $\beta$-suffixes.
Let $\getmemorable{S}{\beta}$ be the set of timers $x_i$ in
$x_1 , \ldots x_{|\beta|}$ whose corresponding timeout event
(of form $\toevent{x_i}$) occurs (as an input) in some $\beta$-suffix.
Let $\getassignment{S}{\beta}$ map each timer $x_i$ in
$\getmemorable{S}{\beta}$ to the unique positive integer to which it
is assigned in the $i$th transition of $\beta$.

We can now defined the Nerode equivalence.

\begin{definition}
  \label{def:bj-nerode}
Let $S$ be a timer language \todobj{Define this},
let $\beta$ and $\beta'$ be canonical untimed behaviors in $S$, 
and let $f : \getmemorable{S}{\beta} \to \getmemorable{S}{\beta'}$
be a bijection
from $\getmemorable{S}{\beta}$ to $\getmemorable{S}{\beta'}$.
Then $\beta$ and $\beta'$ are \emph{equivalent} under $f$, written
$\beta \equiv_{S}^f \beta'$ iff
\[
\gamma \in \suffixbehs{S}{\beta}
\qquad \mbox{iff} \qquad
f(\gamma) \in \suffixbehs{S}{\beta'}
\]
\end{definition}
Intuitively, $\beta \equiv_{S}^f \beta'$ means that $\beta$ and $\beta'$
allow the same suffixes, after renaming
timers assigned in $\beta$ by $f$.
We write $\beta \equiv_{S} \beta'$ to denote that
$\beta \equiv_{S}^f \beta'$ for some
$f : \getmemorable{S}{\beta} \to \getmemorable{S}{\beta'}$.

\begin{theorem}
  \label{thm:bj-nerode}
Let $S$ be a timer language.
Then there exists an MMT with feasible canonical untimed behaviors $S$ iff 
$\equiv_S$ has finitely many equivalence classes (finite index).
\end{theorem}
\begin{proof} 
``$\Leftarrow$'' This direction will produce the following MTT:
\begin{definition}[Constructed MMTautomaton]
\label{def:mmt-construction}
The constructed MMT is the MMT
$\M = (I, O, Q, q_0, \vars, \delta, \lambda, \remap)$, where
\begin{itemize}
\item $Q$ is the set of equivalence classes under $\equiv_S$. We let each
  equivalence class be represented by one of its members, called its
  {\em access string}. In the following, when we write $[\beta]$,
  this denotes that $\beta$ is the access string of the equivalence class $[\beta]$.
\item $q_0 = \emptyword$,
\item $\vars$ maps each equivalence  class $[\beta]\in U$ to
  $\getmemorable{S}{\beta}$,
\item $\lambda$ maps each $[\beta] \in U$ and $i \in \feasibleinputs{\beta}{S}$ to
  the unique $o$ such that $\extend{\beta}{i}{o} \in S$. 
  %% $i$, either in $I$ or of form $\toevent{x_i}$, to the output observed in
  %% reponse to $i$ after $\beta$.
\item $\delta$ maps each $\beta \in U$ and
  $i \in \feasibleinputs{\beta}{S}$ to
  to the unique $\beta' \in U$ such that there is an $f$ with
  $\extend{\beta}{i}{\lambda(\beta,i)} \equiv_{S}^f \beta'$.
\item $\remap$
 maps each $\beta \in U$ and $i \in \feasibleinputs{\beta}{S}$ to
  the function
  $\remap(\beta,i):\getmemorable{S}{\beta'} \mapsto (\getmemorable{S}{\beta} \cup \natplus)$ defined as $\rho \circ f^{-1}$, where $\rho$ maps 
  timer $x_{(|\beta|+1)}$ to
$\getassignment{S}{(\extend{\beta}{i}{\lambda(\beta,i)})}(x_{(|\beta|+1)})$.
\end{itemize}
\end{definition}
\todobj{The rest of this is just a sketch. More care is needed with timers}
``$\Rightarrow$'' Let $\M$ be an MMT, let $S$ be its set of feasible untimed
  behaviors. It follows from \todobj{fix} that $S$ is a timer language.
Suppose that $\beta, \beta' \in S$.
Then there exist (unique) $\delta, \delta$ of $T$ with
$\can{\delta}=\beta$ and $\can{\delta'}=\beta'$.
Suppose $\delta$ and $\delta'$ lead to the same state $q$ of $\M$.
We claim that $\beta \equiv_S \beta'$.
Since $\beta = \can{\delta}$, there exist isomorphisms
$g = g_0 ,\ldots, g_k$  $\beta$ to $\delta$.
Similarly, there exists isomorphisms $g' = g'_0 ,\ldots, g'_l$ from
$\beta'$ to $\delta'$.
Let $f = (g'_l)^{-1} \circ g_k$.
Suppose that, for some untimed behavior $\gamma$, $\beta \cdot \gamma \in S$.
Then there exists an untimed behavior $\zeta$ such that $\delta \cdot \zeta \in T$ and $\can{\delta \cdot \zeta} = \beta \cdot \gamma$.
Moreover, there exists an isomorphism $h = g_k ,\ldots, g_m$ from $\gamma$ to $\zeta$.
Since $\delta \cdot \zeta$ is a feasible behavior in $T$, it follows that
$\delta' \cdot \zeta$ is a feasible behavior in $T$
\todobj{This should follow from the observation that the same paths are possible
  after $\delta$ and $\delta'$ (we have the assumption that all paths are
  feasible}
Thus there exists an untimed behavior $\gamma'$ such that $\can{\delta' \cdot \zeta} = \beta' \cdot \gamma' \in S$.
Moreover, we may construct an isomorphism $h' = g'_l ,\ldots, g'_n$ from $\gamma'$ to $\zeta$.
Then $f = (g'_l)^{-1} \circ g_k \cdots (g'_n)^{-1} \circ g_m$ is an isomorphism from $\gamma$ to $\gamma'$ that extends $f_0$.
We then have $\beta' \cdot f(\gamma) \in S$, as required.
\todobj{To be continued}
\end{proof}


\section{Realizing untimed queries by timed queries}
\todobj{This section has been untouched for a while. Ignore it for now.}

In order to use the above approximated equivalence for learning, we need to
present a procedure for constructing the set
$\apprsuffixbehs{S}{\beta}{V}$ by a sequence of timed membership queries.

We should also construct an automaton from a set of results of such
suffix queries, provided they satisfy some conditions
\todobj{closedness, etc, which we must define}

Let us first present a procedure for answering suffix queries. It is
sufficient to perform them for one generic suffix at a time.
The input to a suffix query is a (feasible) untimed behavior $\beta$ an a
generic untimed input behavior $v$. It returns
$\apprsuffixbehs{S}{\beta}{\set{v}}$. To compute this, it must be determined
\begin{itemize}
\item which instantiations of generic timeouts in $v$ are feasible after $\beta$, and
\item what is the output in response to an input.
\end{itemize}
The second item is trivially answered by observing the output in response to
each relevant input. Note that if a behavior $\beta$ is feasible, then
any extension of $\beta$ by a non-timeout input is trivially feasible, simply
by supplying the input as fast as possible. Thus the only non-trivial part
of answering a suffix query is to determine which instantiations of a generic
timeout in $v$ are feasible. This is easily reduced to the problem of
determining, for a feasible behavior $\beta$, which timeouts of form
$\toevent{x_i}$ with $i \leq |\beta|$ are feasible (note that $\beta$ here in
general denotes an extension of the $\beta$ supplied as prefix in a
suffix query). Let us now describe how this can be done.

In order to determine which behaviors of form $\beta \cdot \toevent{x_i}$
are feasible, we perform a timed membership query, which provide the longest
possible time delay between the $i$th transition in $\beta$ and the time
of any other timeout that can feasibly expire after $\beta$. In order to
maximize this time, we need to know which other timeouts may expire while
performing $\beta \cdot \toevent{x_i}$. Such timeouts are either expiring
as part of $\beta$, or disturb the execution of $\beta \cdot \toevent{x_i}$,
in which case our membership query must know how to prevent them from occuring.

Les us introduce notation. Let $\beta$ be of form
\(
  {i_1/o_1, \rho_1} \ \cdot \ {i_n/o_n, \rho_n}
\)
We are looking for a corresponding timed behavior of form 
\[
\tvals_0 \dtrans{d_1} \tvals'_0 \uttrans{i_1}{o_1}{\rho_1}
\quad \cdots \quad
\dtrans{d_n} \tvals'_n \uttrans{i_n}{o_n}{\rho_n} \tvals_n
\dtrans{d_{n+1}} \tvals'_{n+1} \uttrans{\toevent{x_i}}{o_{n+1}}{\rho_n}
\]
The timed membership queries need to select values
of $\vect{d}{n+1}$, and gradually adds new timers to the word together
with constraints on $\vect{d}{n+1}$ induced by these timers.
Let $\delay{j}{k}$ denote $d_{j+1} + d_{j+2} + \cdots + d_k$.

The algorithm maintains an increasing constraint $\constr$ on delays of form
$\delay jk$. Initially, the constraint contains the conjunct
$\delay jk = D$ whenever $i_k$ is of form $\toevent{x_j}$ and
$\rho_j$ sets $x_j$ to $D$.
Thereafter, the following procedure is performed
  \begin{enumerate}
  \item
    find $\vect d{{n+1}}$ that maximizes $\delay{i}{(n+1)}$ given
    $\constr$ (note that $d_{n+1} = \infty$ is allowed)
  \item
    supply membership query with delay values $\vect d{{n+1}}$.
\begin{itemize}
  \item
    {\bf if} the output is of form $\beta$ followed by $\toevent{x_i}$, then
    report that $\toevent{x_i}$ is feasible and stop
  \item
    {\bf else if} no timeout occurs within $d_{n+1}$ after $\beta$, report
    that $\toevent{x_i}$ is not feasible and stop
  \item
    {\bf else if} another timeout of form $\toevent{x_j}$ occurs before
    the $k$th input, this means that timer $x_j$ is assigned a value
    $D_j$ at the $j$ transition and may expire before the $k$th transition
    in the untimed trace. In order to prevent this from happening,
    add the constraint $\delay jk < D_j$ to $\constr$ and start again
    from step 1.
\end{itemize}
  \end{enumerate}
  Intuitively, the procedure aims to find a timed membership query which maximise
  the time within which $\toevent{x_i}$ is able to expire,
  subject to the constraints under which timers that expire in $\beta$ expire
  as scheduled, and under which any other timers do not expire. These
  other timers may not be known when the procedure starts, and so their
  corresponding constraints are added when they are detected: this happens
  when they expire unexpectectly.

  Note that conjunctions of constraints on delays $\delay{j}{k}$ can be represented as DBM, as is standard for timed automata.




