\section{A Myhill Nerode Theorem for MMTs}
\label{sec:nerode}

\begin{definition}
\label{def:timer language}
A \emph{timer language} over $I$ and $O$ is a nonempty set 
$S$ of feasible untimed behaviors in canonical form over $I$ and $O$ that satisfies the following five properties:
\begin{itemize}
\item
\emph{no initial timers}: $\beta \in S \Rightarrow \Head{\beta} = \emptyset$,
\item
\emph{prefix closed}: $\beta \cdot \gamma \in S \Rightarrow \beta \in S$,
\item
\emph{behavior deterministic}:
$\beta \xrightarrow{i/o_1/\rho_1} X_1 \in S \wedge \beta \xrightarrow{i/o_2/\rho_2} X_2 \in S \Rightarrow o_1 = o_2 \wedge \rho_1 = \rho_2 \wedge X_1 = X_2$,
\item
\emph{input complete}:
$\beta \in S \wedge i \in I \Rightarrow \exists o, \rho, Y : \beta \xrightarrow{i/o/\rho} Y \in S$,
and
\item
\emph{timeout complete}:
$\beta \in S \wedge x \mbox{ expirable after } \beta \Rightarrow
\exists o, \rho, Y: \beta \xrightarrow{\toevent{x}/o/\rho} Y \in S$.
\end{itemize}
\end{definition}

\begin{definition}[Nerode congruence]
\label{def:nerode}
Let $S$ be a timer language.
Suppose $\beta, \beta' \in S$ and $f : \Last{\beta}  \to \Last{\beta'}$ is a bijection.
Then $\beta$ and $\beta'$ are \emph{equivalent under} $f$, written $\beta \equiv^f_S \beta'$, iff  for any untimed behavior $\gamma$:
(a) if $\beta \cdot \gamma \in S$ then there exists an isomorphism $F$ that extends $f$ such that $\beta' \cdot F(\gamma) \in S$, and conversely,
(b) if $\beta' \cdot \gamma \in S$ then there exists an isomorphism $F$ that extends $f^{-1}$ such that $\beta \cdot F(\gamma) \in S$.
We say that $\beta$ and $\beta'$ are \emph{equivalent}, notation $\beta \equiv_S \beta'$ iff they are equivalent under some $f$.
\end{definition}

One easily sees that $\equiv_S$ is an equivalence relation.
%It is also a \emph{right-congruence}, meaning that it satisfies,
%for all $\delta = \beta \xrightarrow{i/o, \rho} Y \in S$ and $\delta' = \beta' \xrightarrow{i/o', \rho'} Y' \in S$,
%\begin{eqnarray*}
%\beta \equiv_S \beta' & \Rightarrow & o = o' \wedge \delta \equiv_S \delta'.
%\end{eqnarray*}
We write $[\beta]$ to denote the equivalence class of $\beta$ w.r.t.\  $\equiv_S$.

If $g : X \to X$ is an injective function and $\beta$ is an untimed behavior, then $\beta [g]$ denotes the untimed behavior obtained
by consistently renaming each occurrence of each variable $x$ by $g(x)$. Using renamings we can give the following characterization
of the Nerode congruence.
Suppose $\beta, \beta' \in S$, $\beta$ contains $m$ inputs, $\beta'$ contains $n$ inputs,
and $f : \Last{\beta}  \to \Last{\beta'}$ is a bijection. 
Then injective function $g_{\beta, \beta', f}: X \to X$ is defined as follows:
\begin{eqnarray*}
g_{\beta, \beta', f} (x_j) & = & \left\{ 
\begin{array}{ll}
f(x_j) & \mbox{if } x_j \in \domof{f}\\
x_{j-m+n} & \mbox{if } j > m
x_j    & \mbox{otherwise }
\end{array}
\right.
\end{eqnarray*}
Then $\beta \equiv^f_S \beta'$ iff for all $\gamma$:
(a) if $\beta \cdot \gamma \in S$ then $\beta' \cdot (\gamma[g_{\beta, \beta', f}]) \in S$, and
(b) if $\beta' \cdot \gamma \in S$ then $\beta \cdot (\gamma[g_{\beta', \beta, f^{-1}}]) \in S$.

\begin{theorem}
\label{Myhill Nerode}
Let $S$ be a timer language.
Then there exists an MMT with feasible untimed behaviors $T$ and $\can{T} = S$ iff 
$\equiv_S$ has finitely many equivalence classes (finite index).
\end{theorem}
\begin{proof} 

``$\Rightarrow$'' Let $\M$ be an MMT, let $T$ be its set of feasible untimed behaviors, and let $S = \can{T}$.
Then it follows from the definitions, Lemma~\ref{feasible plus input is feasible} and Lemma~\ref{expirable} that $S$ is a timer language.
Suppose that $\beta, \beta' \in S$.
Then there exist (unique) $\delta, \delta \in T$ with
$\can{\delta}=\beta$ and $\can{\delta'}=\beta'$.
Suppose $\delta$ and $\delta'$ lead to the same state $q$ of $\M$ and moreover $\Zone{\delta} = \Zone{\delta'}$.
We claim that $\beta \equiv_S \beta'$.
Since $\beta$ and $\delta$ are isomorphic, there exists an isomorphism $g = g_0 ,\ldots, g_k$ from $\beta$ to $\delta$.
Similarly, since $\beta'$ and $\delta'$ are isomorphic, there exists an isomorphism $g' = g'_0 ,\ldots, g'_l$ from $\beta'$ to $\delta'$.
Let $f_0 = (g'_l)^{-1} \circ g_k$.
Suppose that, for some untimed behavior $\gamma$, $\beta \cdot \gamma \in S$.
Then there exists an untimed behavior $\zeta$ such that $\delta \cdot \zeta \in T$ and $\can{\delta \cdot \zeta} = \beta \cdot \gamma$.
Moreover, there exists an isomorphism $h = g_k ,\ldots, g_m$ from $\gamma$ to $\zeta$.
Since $\delta \cdot \zeta$ is a feasible behavior in $T$, it follows by Lemma~\ref{lemma: feasibility concatenation} that
$\delta' \cdot \zeta$ is a feasible behavior in $T$.
Thus there exists an untimed behavior $\gamma'$ such that $\can{\delta' \cdot \zeta} = \beta' \cdot \gamma' \in S$.
Moreover, we may construct an isomorphism $h' = g'_l ,\ldots, g'_n$ from $\gamma'$ to $\zeta$.
Then $f = (g'_l)^{-1} \circ g_k \cdots (g'_n)^{-1} \circ g_m$ is an isomorphism from $\gamma$ to $\gamma'$ that extends $f_0$.
We then have $\beta' \cdot f(\gamma) \in S$, as required.
Clause (b) of Definition~\ref{def:nerode} follows by a symmetric argument.
Since $\M$ only has finitely many states and by Lemma~\ref{lemma finitely many zones} the number of zones of feasible
untimed behaviors of $M$ is finite, it follows that $\equiv_S$ has finite index.

``$\Leftarrow$'' 
Suppose $S$ is a timer language such that $\equiv_S$ has a finite index.
A subset $Q \subseteq S$ is a \emph{basis} for $S$ if it contains a unique representative from
each equivalence class of $\equiv_S$.
%We construct a basis for $S$ using a simple greedy algorithm. Start by setting $Q := \{ \emptyset \}$, where $\emptyset$ is the trivial untimed behavior. Then, as long as an untimed behavior $\beta$ exists that extends a behavior in $Q$ with a single transition and
%is not equivalent to any behavior in $Q$, add it to $Q$.  Since $\equiv_S$ has finite index this algorithm will terminate.
%By construction, the resulting set $Q$ is nonempty, finite and prefix closed.
%By contradiction, we prove that $Q$ contains a representative from each equivalence class.
%Suppose $\beta \in S$ is an untimed behavior that is not equivalent to any behavior in $Q$.
%W.l.o.g.\ we may assume that all proper prefixes of $\beta$ are equivalent to a behavior in $Q$.
%Thus $\beta$ is of the form $\delta \xrightarrow{i/o, \rho} Y$, for some behavior $\delta$ not in $Q$ (otherwise our algorithm would
%have added $\beta$ to $Q$) but equivalent to some $\delta'$ in $Q$.
%Because $S$ is input complete, it contains a behavior $\beta' = \delta' \xrightarrow{i/o', \rho'} Y'$.
%Since $\equiv_S$ is a right-congruence, it follows that $\beta \equiv_S \beta'$.
%Since $\beta$ is not equivalent to any behavior in $Q$, the same holds for $\beta'$.
%This is a contradiction, because we have found a behavior $\beta'$ that extends a behavior in $Q$ with a single transition,
%is not equivalent to any behavior in $Q$, but nevertheless has not been added to $Q$.
%
Since $\equiv_S$ has finite index, $S$ has a finite basis $Q$.
Now MMT $\M = (I, O \cup \{ \perp \}, Q, \beta_0, \vars, \delta, \lambda, \remap)$ is defined as follows:
\begin{itemize}
\item
$\beta_0$ is the unique element of $Q$ such that $\beta_0 \equiv_S \emptyset$.
\item
$\varsof{\beta} = \Last{\beta}$.
\item
Let $\beta \in Q$ and $i \in I$. Then, since $S$ is input complete and behavior deterministic, there are unique
$o$, $\rho$ and $Y$ such that $\zeta = \beta \xrightarrow{i/o/\rho} Y \in S$.
Let $\zeta'$ be the unique element of $Q$ with $\zeta' \equiv_S \zeta$, and
let $f_0$ be a matching for $\zeta' \equiv_S \zeta$.
Then $\delta(\beta,i) = \zeta'$, $\lambda(\beta,i) = o$, and $\remap(\beta,i) = \rho  \circ f_0$.
\item
Assume $\beta \in Q$ and $x\in \Last{\beta}$ expirable after $\beta$. 
Then, since $S$ is  timeout complete and behavior deterministic, there exist unique
$o$, $\rho$ and $Y$ such that $\zeta = \beta \xrightarrow{\toevent{x}/o/\rho} Y \in S$.
Let $\zeta'$ be the unique element of $Q$ with $\zeta' \equiv_S \zeta$, and
let $f_0$ be a matching for $\zeta' \equiv_S \zeta$.
Then $\delta(\beta,\toevent{x}) = \zeta'$, $\lambda(\beta,\toevent{x}) = o$, and $\remap(\beta,\toevent{x}) = \rho \circ f_0$.
\item
Assume $\beta \in Q$ and $x \in \Last{\beta}$ not expirable after $\beta$.
We define $\delta(\beta,\toevent{x}) = \beta$, $\lambda(\beta,\toevent{x}) = \perp$, and $\remap(\beta,\toevent{x}) = \rho_0$, 
where $\perp$ is a special, fresh output event and $\domof{\rho_0} = \emptyset$.
\end{itemize}
It is routine to verify that $\M$ is a well-defined MMT whose set of feasible untimed behaviors is isomorphic to $S$.
\end{proof}
