\newcommand{\Zone}[1]{\mathsf{Zone}({#1})}

\subsection{Zones and constraints}

In order to implement an untimed teacher using a timed teacher, we need some basic facts and algorithms to
analyse and manipulate untimed behaviors and timed words that we will discuss in this subsection.

%It is not difficult to translate an MMTs to a timed automaton \cite{AD94,BengtssonY03} that accepts the same timed words.
%Through such a translation, verification and analysis tools for timed automata, such as Uppaal \cite{Uppaal4.0}
%become available for MMTs.
%MMTs are less expressive than timed automata. MMTs, for instance, do not contain time deadlocks or zeno loops that
%may prevent time from progressing.

MMTs have infinitely (in fact even uncountably) many configurations and it is thus necessary to use symbolic representations
for reachability analysis. Given an untimed behavior $\beta$ we want to compute its effect on a set of valuations. 
If $\beta =X_0$ and $K \subseteq\Vals{X_0}$ then we define $\Post_{\beta}(K) = K$.
If $\beta = X_0 \xrightarrow{i_1/o_1, \rho_1} X_1$ and
$K \subseteq\Vals{X_0}$ then $\Post_{\beta}(K)$ is equal to the set of all valuations $\tvals'' \in\Vals{X_1}$ such that
\[
  \exists \tvals \in K \exists d >0 \exists \tvals' \in\Vals{X_0} :
 \tvals \xrightarrow{d} \tvals' \xrightarrow{i_1/o_1, \rho_1} \tvals'' .
\]
The definition of $\Post_{\beta}(K)$ is extended inductively to arbitrary $\beta$ by 
$\Post_{\gamma \cdot \gamma'}(K) = \Post_{\gamma'} (\Post_{\gamma}(K))$, where $\gamma \cdot \gamma'$ is the decomposition of
$\beta$ into an untimed behavior $\gamma$ of length one and an untimed behavior $\gamma'$.
We write $\Zone{\beta}$ as abbreviation for $\Post_{\beta}(\Vals{\Head{\beta}})$.
%
For any untimed behavior $\beta$, $\Zone{\beta}$ can be symbolically represented and computed using a Difference Bound Matrices (DBMs)
 \cite{Di89}, as the transition relations $\xrightarrow{d}$ and $\xrightarrow{i_1/o_1, \rho_1}$ can be decomposed 
into elementary operations on DBMs such as reset, conjunction, and delay successors \cite{BengtssonY03}.

%It is easy to see that an untimed behavior $\beta$ is feasible iff $\Zone{\beta} \neq \emptyset$.
%Since emptiness of DBMs is decidable, this allows us to compute whether or not $\beta$ is feasible.

The next technical lemma's are needed further on:

\begin{lemma}
\label{lemma: feasibility concatenation}
Suppose $\beta, \beta'$ are untimed behaviors such that
$\Zone{\beta} = \Zone{\beta'}$. Let $\gamma$ be any untimed behavior.
Then $\beta \cdot \gamma$ is feasible iff $\beta' \cdot \gamma$ is feasible.
\end{lemma}

\begin{lemma}
\label{lemma finitely many zones}
$\{ \Zone{\beta} \mid \beta \mbox{ feasible untimed behavior of } \M \}$ is finite.
\end{lemma}
\iflong
\begin{proof}
All the sets $\Zone{\beta}$ can be represented using DBMs. An MMT only has a finite number of timers that can only be set to a finite number of integer values. Since in an MMT the values of timers can only decrease, only finitely many numbers may
appear in the DBM's that represent the sets $\Zone{\beta}$. Thus all the sets $\Zone{\beta}$ can be represented by a finite
collection of DBMs.
\end{proof}
\fi

Let $\beta$ be a feasible untimed behavior and let $x \in X$ be a timer. Then we say that $x$ is \emph{expirable} after $\beta$
if there exists a valuation in $\Zone{\beta}$ in which $x$ is minimal.

\begin{lemma}
\label{expirable}
Suppose $\beta$ is a feasible untimed behavior with $\Last{\beta} = Y$ and $Y \xrightarrow{\toevent{x}/o, \rho} Y'$ is an untimed behavior.
Then $x$ is expirable after $\beta$ iff $\beta \xrightarrow{\toevent{x}/o, \rho} Y'$ is feasible.
\end{lemma}

Suppose that $\beta$ is an untimed behavior in canonical form:
\begin{eqnarray*}
\beta & = & \emptyset \xrightarrow{i_1/o_1, \rho_1} X_1  \cdots X_{k-1} \xrightarrow{i_k/o_k, \rho_k} X_k.
\end{eqnarray*}
We associate to $\beta$ a set of constraints with real-valued variables $t_1 ,\ldots t_k$ that denote the time of
occurrence of events $i_1 ,\ldots, i_k$ respectively. We add the following constraints to this set $\Constraints{\beta}$:
\begin{itemize}
\item
$0 < t_1 \leq d_{\max}$,
\item
for each index $j < k$:  $0 <  t_{j+1} - t_j \leq d_{\max}$,
\item
for each timeout event $i_j = \toevent{x_l}$: $t_j = t_l + \rho_j(x_l)$,
\todobj{This is a typo? should be $\rho_l(x_l)$?}
\item
for each clock $x_l$ that is started but does not timeout: $t_j \leq t_l + \rho_j(x_l)$,
where $j$ is the largest index such that $x_l \in X_j$, and
\item
for each pair of distinct indices $j$ and $l$ with $i_j, i_l \in I$: $\Frac{t_j} \neq \Frac{t_l}$ 
(to express that the fractional parts of $t_j$ and $t_l$ are different).
\end{itemize}
Note that $\beta$ is feasible iff the set of constraints $\Constraints{\beta}$ is satisfiable.
We may use an SMT solver to decide whether $\Constraints{\beta}$ is satisfiable and to compute\footnote{If an SMT solver
does not support fractions, we may compute a solution without these constraints and then use the wiggling approach
of Section~\ref{section untimed semantics} to ensure that the fractional times of inputs are different.} a solution.
\todofv{Add example}

\todofv{Move next lemma to earlier section}

\begin{lemma}
\label{not timed}
Suppose $\M, \N$ are MMTs with $\M \not\approx_{\mathit{timed}} \N$.
Then there exists a transparent timed word $w$ of $\M$ that is not a timed word of $\N$.
\end{lemma}

%\marginpar{What is the complexity of deciding eg timed equivalence or reachability of MMTs?}
